Enhancing Accelerometer-based Activity Recognition with Capacitive Proximity Sensing

### 2012-06-22 Paper/Paper.tex

\maketitle
\begin{abstract}
Activity recognition solely with an accelerometer is a common investigated research topic and enables the detection of basic activities like sitting, walking or standing. Recent work in this area adds different sensing modalities to the inertial data to collect more information of the users environment and to boost activity recognition. This work presents a sensor prototype which consists of an accelerometer and a capacitive proximity sensor that senses the users activity based on the combined sensor values. We show that our proposed approach improves the recognition rate significantly for detecting activities of daily living (e.g. sitting, preparing dinner, eating, drinking, etc.) by adding capacitive proximity sensor data to the inertial data.
\keywords{ambient assisted living, capacitive proximity sensors, activity recognition, user context}
\end{abstract}

@@

% This paper is structured as follows: We give a short insight in related work in chapter 2, followed by describing the hardware set-up used for this study in chapter 3. The experiment and evaluation will be described in chapter 4, showing how data has been obtained and how activity recognition is being improved. This work is concluded in chapter 5 with an outlook to future work.

The remainder of this paper is  is structured as follows: First, Section \ref{sect:related} is dedicated to related work work in activity recognition, in particular considering approaches with multiple sensor modalities. Section \ref{sect:hardware} will present the hardware based on which a wrist-worn sensor prototype was built, fusing together an inertial data logger and a capacitive sensor integrated in a wristband. The experimental setup with the scripted scenario of daily activities and the activity recognition evaluation results, showing the performance boost of the capacitive sensor unit, are given in Section \ref{sect:experiment}. The paper is wrapped up with a conclusions section enumerating the findings of the paper, as well as pointing out future research potential.

@@

\subsection{Evaluation Results}

@@

\begin{figure}
	\centering
		\includegraphics[width=1.00\textwidth]{../Auswertung/images/tobias_1.pdf}
	\caption{While entering the door, the wristband approaches the door knob twice, while opening and closing. Regarding the capacitive proximity sensor, this situation can be observed at the beginning and at the end of the activity.}
	\label{fig:tobias_1}
\end{figure}

The opening door activity has very poor recongition rates without the including the proximity sensor. The average f-measure could be increased from of 34.2 \% to 46.8 \%. A plot of the activity is given in Figure \ref{fig:tobias_1}. The capacitive proximity sensor shows two approaches to the door knob, one for opening the door (2 s) and one for closing the door (9 s). The acceleration sensor captures relevant data in the time in which the person moves into the room and the hand changes from the outer to the inner door knob (5 - 7 s). The recorded data for this activity shows strong correlations between all test persons.

\begin{figure}
	\centering
		\includegraphics[width=1.00\textwidth]{../Auswertung/images/eugen_2.pdf}
	\caption{Example of a ``sitting'' activity in which user changed his position very frequently. Throughout the whole activity the values of the proximity sensor stay constant due to the close proximity to the couch's fabric.}
	\label{fig:eugen_2}
\end{figure}

@@

In the following, the test persons were supposed to lay down on the couch. Again, there were great variations in performing this activity between all participants. For example, some of them crossed their hands under their head. For that class, the average f-measure could only be increased by 2.4 \%, from 82.1 to 84.5 \%. The participants were also asked to put food and dishes from a shelf and a locker on a table. The capacitive proximity sensor captured the proximity to the shelf and to the table. The overall f-measure for this activity is quite low, but improved by 6.7 \% from 50.4 \% to 57.1\%.

\begin{figure}
	\centering
		\includegraphics[width=\textwidth]{../Auswertung/images/eugen_5.pdf}
	\caption{An example of the ``preparing bread'' activity, where the participants had to put marmalade on a toast. The proximity sensor indicates the closeness to the table, while the acceleration sensor shows reoccurring movements.}
	\label{fig:eugen_5}
\end{figure}

Figure \ref{fig:eugen_5} shows an example instance of preparing a bread with marmelade. It is notable that the acceleration data does not seem to provide any characteristic patterns, while the proximity sensor indicates a table, plate, or other objects in immediate distance. This activity showed a high improvement in the average f-measure by 10 \%, from 49.0 \% to 59.8 \%, when the capacitive proximity sensor is taken into account.

\begin{figure}
	\centering
		\includegraphics[width=\textwidth]{../Auswertung/images/eugen_6.pdf}
	\caption{An eating person who took 5 bites from the bread. After each bite, the hand is placed on the table.}
	\label{fig:eugen_6}
\end{figure}

@@

\begin{figure}
	\centering
		\includegraphics[width=\textwidth]{../Auswertung/images/eugen_7.pdf}
	\caption{An example of the ``drinking'' activity of a test person. The participant first pours some water into the glass and then takes three drams of water. After each dram, he returns his arm to the table which can be observed in the characteristic patterns of the proximity sensor.}
	\label{fig:eugen_7}
\end{figure}

The activity drinking water is depicted in Figure \ref{fig:eugen_7}. The participant took five nips from the glass, while leaving the hands lying on the table in between. These motions can be easily detected in the acceleration as well as proximity data. The accelerometer data shows that there periodic up- and down-movements while the capacitive proximity sensor delivers data that is associated to the proximity of the table.

\begin{figure}
	\centering
		\includegraphics[width=1.00\textwidth]{../Auswertung/images/marko_8.pdf}
	\caption{An examplary instance of the class ``walking''. The acceleration sensor and the proximity sensor show periodic reoccuring patterns that are related to the pendulum-like arm movement and the proximity to the person's body during those movements.}
	\label{fig:marko_8}
\end{figure}

Regarding the walking activity one can identify periodic changes in the acceleration as well as in the measured capacitance, illustrated in Figure \ref{fig:marko_8}. While walking, the capacitance between the wristband and the leg increases when the wristband is located close to the body and decrease when the wristband moves away.

\begin{figure}
	\centering
		\includegraphics[width=1.00\textwidth]{../Auswertung/images/eugen_9.pdf}
	\caption{During the sleeping activity the data from both sensors remains constant. One draw conclusions about the coverage of the arm with either couchons, blankets or the proximity to the mattress.}
	\label{fig:eugen_9}
\end{figure}

@@

\begin{table}
	\caption{Activity recognition evaluation revealing the positive impact of the capacitive proximity sensor. Here, we are comparing SVM classification presented as confusion matrices for an exemplary user, without the proximity data on the left, and with the proximity data on the right.}

@@

Table \ref{tab:ConfusionMatrixForAnExemplaryUser}

>>>>

\maketitle
\begin{abstract}
Activity recognition solely with an accelerometer is a common investigated research topic and enables the detection of basic activities like sit, walk or stand. Recent work in this area adds different sensing modalities to the inertial data to collect more information of the users environment and to boost activity recognition.  We present a sensor set-up that consists of an accelerometer and a capacitive proximity sensor that senses the users activity based on the combined sensor values. We show that our proposed approach improves the recognition rate significantly for detecting activities of daily living (e.g. open door, prepare dinner, etc.) by adding capacitive proximity sensor data to the inertial data.
\keywords{ambient assisted living, capacitive proximity sensors, activity recognition, user context}
\end{abstract}

@@

This paper is structured as follows: We give a short insight in related work in chapter 2, followed by describing the hardware set-up used for this study in chapter 3. The experiment and evaluation will be described in chapter 4, showing how data has been obtained and how activity recognition is being improved. This work is concluded in chapter 5 with an outlook to future work.

The remainder of this paper is  is structured as follows: First, Section \ref{sect:related} is dedicated to related work work in activity recognition, in particular considering approaches with multiple sensor modalities. Section \ref{sect:hardware} will present the hardware based on which a wrist-worn sensor prototype was built, fusing together an inertial data logger and a capacitive sensor integrated in a wristband. The experimental setup, the scripted scenario with daily activities and the activity recognition evaluation results, showing the performance boost of the capacitive sensor unit, are given in Section \ref{sect:experiment}. The paper is wrapped up with a conclusions section enumerating the findings of the paper, as well as pointing out future research potential.

@@

@@

\begin{figure}[htbp]
	\centering
		\includegraphics[width=1.00\textwidth]{../Auswertung/images/tobias_1.pdf}
	\caption{While entering the door, the wristband approaches the door knob twice, while opening and closing. Regarding the capacitive proximity sensor, this situation can be observed at the beginning and at the end of the activity.}
	\label{fig:tobias_1}
\end{figure}

The opening door activity has very poor recongition rates without the including the proximity sensor. The average f-measure could be increased from of 34.2 \% to 46.8 \%. A plot of the activity is given in Figure \ref{fig:tobias_1}. The capacitive proximity sensor shows two approaches to the door knob, one for opening the door (2 s) and one for closing the door (9 s). The acceleration sensor captures relevant data in the time in which the person moves into the room and the hand changes from the outer to the inner door knob (5 - 7 s). The recorded data for this activity shows strong correlations between all test persons. The confusion matrices show that the open door activity was often mixed up with the sitting activity because of the horizontal orientation of the hand. Due to the usage of the capacitive sensor, the recall for that class and confusion with the sitting activity improved.

\begin{figure}[htbp]
	\centering
		\includegraphics[width=1.00\textwidth]{../Auswertung/images/eugen_2.pdf}
	\caption{Example of a ``sitting'' activity in which user changed his position very frequently. Throughout the whole activity the values of the proximity sensor stay constant due to the close proximity to the couch's fabric.}
	\label{fig:eugen_2}
\end{figure}

@@

In the following, the test persons were supposed to lay down on the couch. Again, there were great variations in performing this activity between all participants. For example, some of them crossed their hands under their head. For that class, the average f-measure could only be increased by 2.4 \%, from 82.1 to 84.5 \%. Considering some test persons, this class was confused with the make bread activity and could be improved by the proximity sensor.

The participants were also asked to put food and dishes from a shelf and a locker on a table. The capacitive proximity sensor captured the proximity to the shelf and to the table. The overall f-measure for this activity is rather low, but improved by 6.7 \% from 50.4 \% to 57.1\%. Reasons???

\begin{figure}[htpb]
	\centering
		\includegraphics[width=\textwidth]{../Auswertung/images/eugen_5.pdf}
	\caption{An example of the ``preparing bread'' activity, where the participants had to put marmalade on a toast. The proximity sensor indicates the closeness to the table, while the acceleration sensor shows reoccurring movements.}
	\label{fig:eugen_5}
\end{figure}

Figure \ref{fig:eugen_5} shows an example instance of preparing a bread with marmelade. It is notable that the acceleration data does not seem to provide any characteristic patterns, while the proximity sensor indicates a table, plate, or other objects in immediate distance. This activity showed a high improvement in the average f-measure by 10 \%, from 49.0 \% to 59.8 \%, when the data delivered by the capacitive proximity sensor is taken into account. The making bread class was often confused with the sitting class. Using the new input modality, this confusion could be improved.

\begin{figure}[htpb]
	\centering
		\includegraphics[width=\textwidth]{../Auswertung/images/eugen_6.pdf}
	\caption{An eating person who took 5 bites from the bread. After each bite, the hand is placed on the table.}
	\label{fig:eugen_6}
\end{figure}

@@

\begin{figure}[htpb]
	\centering
		\includegraphics[width=\textwidth]{../Auswertung/images/eugen_7.pdf}
	\caption{An example of the ``drinking'' activity of a test person. The participant first pours some water into the glass and then takes three drams of water. After each dram, he returns his arm to the table which can be observed in the characteristic patterns of the proximity sensor.}
	\label{fig:eugen_7}
\end{figure}

The activity drinking water is depicted in Figure \ref{fig:eugen_7}. The participant took a few nips from the glass, while leaving the hands lying on the table in between. These motions can be easily detected in the acceleration as well as proximity data. The accelerometer data shows that there periodic up- and down-movements while the capacitive proximity sensor delivers data that is associated to the proximity of the table.

\begin{figure}[htbp]
	\centering
		\includegraphics[width=1.00\textwidth]{../Auswertung/images/marko_8.pdf}
	\caption{An examplary instance of the class ``walking''. The acceleration sensor and the proximity sensor show periodic reoccuring patterns that are related to the pendulum-like arm movement and the proximity to the person's body during those movements.}
	\label{fig:marko_8}
\end{figure}

Regarding the walking activity one can identify periodic changes in the acceleration as well as in the measured capacitance, illustrated in Figure \ref{fig:marko_8}. While walking, the capacitance between the wristband and the leg increases when the wristband is located close to the body and decrease when the wristband moves away. There were problems in distinguishing this activity from get things that could be bettered by using the new input modality.

\begin{figure}[htbp]
	\centering
		\includegraphics[width=1.00\textwidth]{../Auswertung/images/eugen_9.pdf}
	\caption{During the sleeping activity the data from both sensors remains constant. One draw conclusions about the coverage of the arm with either couchons, blankets or the proximity to the mattress.}
	\label{fig:eugen_9}
\end{figure}

@@

\begin{table}[htbp]
	\centering	\begin{tabular}{p{0.48cm}p{0.48cm}p{0.48cm}p{0.48cm}p{0.48cm}p{0.48cm}p{0.48cm}p{0.48cm}p{0.48cm}|p{0.48cm}p{0.48cm}p{0.48cm}p{0.48cm}p{0.48cm}p{0.48cm}p{0.48cm}p{0.48cm}p{0.48cm}|p{2.2cm}}
a & b & c & d & e & f & g & h & i & a & b & c & d & e & f & g & h & i & $\leftarrow$ classified as\\
\hline
10 & 3 & 0 & 2 & 0 & 0 & 0 & 0 & 0 		& 9 & 5 & 0 & 0 & 1 & 0 & 0 & 0 & 0 & a = open door\\
\hline
2 & 259 & 4 & 2 & 4 & 2 & 4 & 0 & 0 	& 1 & 248 & 5 & 6 & 4 & 2 & 5 & 0 & 0 & b = sit\\
\hline
0 & 5 & 213 & 2 & 2 & 1 & 8 & 0 & 4 	& 0 & 3 & 212 & 3 & 2 & 1 & 6 & 0 & 8 & c = lying\\
\hline
3 & 5 & 2 & 52 & 9 & 3 & 1 & 1 & 1 		& 0 & 8 & 2 & 44 & 7 & 8 & 1 & 1 & 3 & d = get things\\
\hline
1 & 9 & 3 & 13 & 46 & 6 & 3 & 0 & 0 	& 1 & 7 & 1 & 11 & 37 & 9 & 2 & 0 & 10 & e = make bread\\
\hline
0 & 5 & 1 & 1 & 5 & 62 & 5 & 0 & 1 		& 0 & 3 & 2 & 3 & 8 & 58 & 5 & 0 & 1 & f = eat\\
\hline
0 & 8 & 4 & 2 & 1 & 5 & 68 & 0 & 0 		& 0 & 8 & 3 & 0 & 6 & 4 & 67 & 0 & 1 & g = drink\\
\hline
1 & 0 & 0 & 1 & 0 & 0 & 0 & 12 & 0 		& 1 & 0 & 0 & 1 & 0 & 0 & 0 & 13 & 0 & h = walk\\
\hline
1 & 0 & 0 & 7 & 0 & 0 & 0 & 0 & 325 	& 0 & 1 & 2 & 10 & 6 & 1 & 1 & 0 & 312 & i = sleep\\
		\end{tabular}
	\caption{Activity recognition evaluation revealing the positive impact of the capacitive proximity sensor. Here, we are comparing SVM classification presented as confusion matrices for an exemplary user, without the proximity data on the left, and with the proximity data on the right.}

@@

Table \ref{tab:confmatrix}
